---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# readsdr

<!-- badges: start -->
<!-- badges: end -->

## Overview

The goal of readsdr is to bridge the design capabilities from specialised
System Dynamics software with the powerful numerical tools offered by R 
libraries. The package accomplishes this goal by parsing .xmile files 
([Vensim](https://vensim.com/) and [Stella](https://www.iseesystems.com/) models) 
into R objects to construct [networks](http://igraph.org) (graph theory), ODE 
functions for [deSolve](http://desolve.r-forge.r-project.org/) and 
[Stan](https://mc-stan.org/) .

## Installation

<!-- You can install the released version of readsdr from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("readsdr") -->
<!-- ``` -->

<!-- And the development version from [GitHub](https://github.com/) with: -->
You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jandraor/readsdr")
```

## Usage

```{r example}
library(readsdr)
filepath <- system.file("models/", "SIR.stmx", package = "readsdr")
mdl      <- read_xmile(filepath) 
summary(mdl)
```

For reading Vensim models, they must be exported as .xmile. 


`vignette("readsdr")` gives more detail on how to use the package.


<!-- ## Acknowledgments -->

<!-- Thanks to: -->
