---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# readsdr

<!-- badges: start -->
[![Codecov test
coverage](https://codecov.io/gh/jandraor/readsdr/branch/master/graph/badge.svg)](https://codecov.io/gh/jandraor/readsdr?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/readsdr)](https://CRAN.R-project.org/package=readsdr)
[![R-CMD-check](https://github.com/jandraor/readsdr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jandraor/readsdr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Overview

The goal of readsdr is to bridge the design capabilities from specialised
System Dynamics software with the powerful numerical tools offered by R 
libraries. The package accomplishes this goal by parsing .xmile files 
(Vensim and [Stella](https://www.iseesystems.com/) models) into R objects to
construct [networks](https://igraph.org) (graph theory), ODE functions for 
[deSolve](http://desolve.r-forge.r-project.org/) and 
[Stan](https://mc-stan.org/).

## Installation

You can install the released version of readsdr from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("readsdr")
```

And the development version from [GitHub](https://github.com/) with:


``` r
# install.packages("devtools")
devtools::install_github("jandraor/readsdr")
```

## Usage

```{r example}
library(readsdr)
filepath <- system.file("models/", "SIR.stmx", package = "readsdr")
mdl      <- read_xmile(filepath) 
summary(mdl)
```

For reading Vensim models, they must be exported as .xmile. 

For information on how to use this package, see the [vignette](https://CRAN.R-project.org/package=readsdr/vignettes/Introduction_to_readsdr.html)
or type `vignette("Introduction_to_readsdr")` in R.

## Supported builtins

* **Stella**
    + Comparison operators (`=`, `<>`)
    + Logical operators (`AND`, `OR`, `NOT`) 
    + `If Else Then`
    + Inequality operators (`<`, `>`)
    + `Pulse` ^1^
    + `Step` ^1^
    + Endogenous table functions
    + Smoothing functions^2^: `SMTH1`, `SMTH3`, `SMTHN`
    + Delay functions: `DELAYN`
    + Limited support to *uni-dimensional* arrays. *Apply all* translation is restricted to numeric values.
    + Math functions: `ABS`, `SQRT`
    + Stats functions: `NORMAL`^3^
* **Vensim**
    + Comparison operators (`=`, `<>`)
    + Logical operators (`:AND:`, `:OR:`, `:NOT:`) 
    + `IF_THEN_ELSE`
    + Inequality operators (`<`, `>`) 
    + `Pulse` ^1^
    + `Pulse Train` ^1^
    + `Step` ^1^
    + Endogenous table functions
    + Smoothing functions^2^: `SMOOTH`, `SMOOTH3`, `SMOOTH3I`, `SMOOTHI` 
    + Delay functions: `DELAY N`
    + Math functions: `ABS`, `SQRT` 
    + Limited support to *bi-dimensional* arrays.
    + Stats functions: `RANDOM NORMAL`^3^
    
^1^ Restricted to Euler integration.

^2^ These functions cannot be part of more complex mathematical expressions. That is, the auxiliary variable must only contain one smoothing function and **nothing else**.

^3^ Seed is ignored.

## Notes

* *uniflow* and *non-negative* stock features from _Stella_ are **not** supported.

* No built-in is supported for translations to _Stan_ code.

* Modules from _Stella_ are **not** supported.


## Applications

This package has been instrumental in the following works:

* [Andrade & Duggan (2023)](https://doi.org/10.1098/rsos.230515). _Anchoring the mean generation time in the SEIR to mitigate biases in_ $\Re_0$ _estimates due to uncertainty in the distribution of the epidemiological delays_. **Royal Society Open Science**.

* [Andrade & Duggan (2022)](https://doi.org/10.1371/journal.pcbi.1010206). _Inferring the effective reproductive number from deterministic and semi-deterministic compartmental models using incidence and mobility data_. **PLOS Computational Biology**.

* [Andrade & Duggan (2021)](https://doi.org/10.1002/sdr.1693). _A Bayesian approach to calibrate system dynamics models using Hamiltonian Monte Carlo_. **System Dynamics Review**.

* [Andrade & Duggan (2020)](https://doi.org/10.1016/j.epidem.2020.100415). _An evaluation of Hamiltonian Monte Carlo performance to calibrate age-structured compartmental SEIR models to incidence data_. **Epidemics**.





## Acknowledgments

Thanks to:

* [Rogelio Oliva](http://people.tamu.edu/~roliva/) for advocating the need to
  create an open-source tool for the System Dynamics community and push forward 
  this endeavour.
* Sergey Naumov for his ideas to implement several functions in this package.
* [Jim Duggan](http://www.nuigalway.ie/our-research/people/engineering-and-informatics/jamesduggan/) 
  from whom I borrowed ideas to implement this package.

## References

[Duggan, J. (2016). *System Dynamics Modeling with R*. Springer.](https://www.springer.com/us/book/9783319340418)


