---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# readsdr

<!-- badges: start -->
[![Codecov test
coverage](https://codecov.io/gh/jandraor/readsdr/branch/master/graph/badge.svg)](https://codecov.io/gh/jandraor/readsdr?branch=master)
[![R build status](https://github.com/jandraor/readsdr/workflows/R-CMD-check/badge.svg)](https://github.com/jandraor/readsdr/actions)
<!-- badges: end -->

## Overview

The goal of readsdr is to bridge the design capabilities from specialised
System Dynamics software with the powerful numerical tools offered by R 
libraries. The package accomplishes this goal by parsing .xmile files 
([Vensim](https://vensim.com/) and [Stella](https://www.iseesystems.com/) models) 
into R objects to construct [networks](http://igraph.org) (graph theory), ODE 
functions for [deSolve](http://desolve.r-forge.r-project.org/) and 
[Stan](https://mc-stan.org/) .

## Installation

<!-- You can install the released version of readsdr from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("readsdr") -->
<!-- ``` -->

<!-- And the development version from [GitHub](https://github.com/) with: -->
You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jandraor/readsdr")
```

## Usage

```{r example}
library(readsdr)
filepath <- system.file("models/", "SIR.stmx", package = "readsdr")
mdl      <- read_xmile(filepath) 
summary(mdl)
```

For reading Vensim models, they must be exported as .xmile. 


`vignette("Introduction_to_readsdr")` gives more detail on how to use the package.

## Supported builtins

* **Stella**
    + Comparison operators (`=`, `<>`)
    + Logical operators (`AND`, `OR`, `NOT`) 
    + `If Else Then`
    + Inequality operators (`<`, `>`)
    + `Pulse` ^1^
    + `Step` ^1^
    + Endogenous table functions
* **Vensim**
    + Comparison operators (`=`, `<>`)
    + Logical operators (`:AND:`, `:OR:`, `:NOT:`) 
    + `IF_THEN_ELSE`
    + Inequality operators (`<`, `>`) 
    + `Pulse` ^1^
    + `Pulse Train` ^1^
    + `Step` ^1^
    + Endogenous table functions
    
^1^ Restricted to Euler integration.

## Notes

* *uniflow* and *non-negative* stock features from Stella are **not** supported.

* No built-in is supported for translations to Stan code.

## Acknowledgments

Thanks to:

* [Rogelio Oliva](http://people.tamu.edu/~roliva/) for advocating the need to
  create an open-source tool for the System Dynamics community and push forward 
  this endeavour.
* Sergey Naumov for his ideas to implement several functions in this package.
* [Jim Duggan](http://www.nuigalway.ie/our-research/people/engineering-and-informatics/jamesduggan/) 
  from whom I borrowed ideas to implement this package.

## References

[Duggan, J. (2016). *System Dynamics Modeling with R*. Springer.](http://www.springer.com/us/book/9783319340418)


