---
title: "Introduction to readsdr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{readsdr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Model: SIR
----------

The SIR model is an epidemiological model, due to [Kermack and McKendrick](https://royalsocietypublishing.org/doi/pdf/10.1098/rspa.1927.0118), 
that computes the theoretical number of people infected with a contagious 
illness in a closed population over time. The SIR models the flows of people 
between three states:

Susceptible (S) 
  : number of individuals who are not infected but could become infected.
  
Infected (I) 
  : number of individuals who have the disease and can transmit it to the 
  susceptibles. 

Recovered (R) 
  : number of individuals who have recovered from the disease and are immune 
  from getting it again.
  

The model assumes a time scale short enough that births and deaths can be 
neglected.

The SIR model is used where individuals infect each other directly. Contact 
between people is also modeled to be random.

The rate that people become infected is proportional to the number of people who
are infected, and the number of people who are susceptible. If there are lots of
people infected, the chances of a susceptible coming into contact with someone 
who is infected is high. Likewise, if there are very few people who are 
susceptible, the chances of a susceptible coming into contact with an infected 
is lower (since most of the contact would be between either infected or 
recovered). In mathematical notation, the model is described by the following
equations:

\begin{equation}
  \frac{dS}{dt} = -\beta SI
\end{equation}

\begin{equation}
  \frac{dI}{dt} = \beta SI - \frac{I}{rd}
\end{equation}

\begin{equation}
  \frac{dR}{dt} = \frac{I}{rd}
\end{equation}

\begin{equation}
  \beta = \frac{e}{n}
\end{equation}

\begin{equation}
  e = c \times i
\end{equation}

$n$ denotes the population size; $\beta$ the per capita rate at which two 
specific individuals come into effective contact per unit time; $e$ the 
effective contacts per infected individual; $c$ the contacts per person per 
unit time; and $i$ the infectivity. 

There exists several ways to implement this model. On one side of the spectrum,
we find specialised software such as [Vensim](https://vensim.com/) and 
[Stella](https://www.iseesystems.com/store/products/stella-architect.aspx) that 
offer friendly graphical user interfaces to seamlessly design complex models, 
emphasising on a systems perspective. _Figure 1_ presents the implementation of
the SIR model in Stella.


```{r, echo=FALSE, fig.cap="Figure 1. SIR", out.width = '80%'}
filepath <- system.file("models/", "SIR_diagram.png", package = "readsdr")
knitr::include_graphics(filepath)
```



On the other side, we can implement the model in flexible and powerful 
statistical environments such as R that offer innumerable tools for numerical 
analysis and data visualisation. Specifically, models can be implemented with 
the [deSolve](https://www.jstatsoft.org/article/view/v033i09) library (see 
[Duggan](https://www.springer.com/gp/book/9783319340418) for more details).
This alternative requires the user to type all model equations in computational
order. For large models, this task can be cumbersome and may lead to incorrect
implementations.

In order to bridge these two approaches, the package **readsdr** fills the gap
by automating the translation from Vensim and Stella models to R. Such a process
is achieved by the function **read_xmile**.


```{r}
library(readsdr)

filepath <- system.file("models/", "SIR.stmx", package = "readsdr")
mdl      <- read_xmile(filepath) 
```

_read_xmile_ returns a list of three elements. The element _description_ 
contains the simulation parameters and the model variables as lists. 

```{r}
description <- mdl$description

model_summary <- data.frame(n_stocks    = length(description$levels),
                            n_variables = length(description$variables),
                            n_consts    = length(description$constants))
print(model_summary)
```

To simulate a model with deSolve, we use the function _ode_. This routine
takes as arguments, a vector with the model's stocks, the simulation time,
the equations wrapped in a function, the model's constants and the integration
method. Indeed, the second element from _read_xmile_ is a list with all these
inputs but the integration method. If this is the only element of interest,
readsdr provides **xmile_to_deSolve**. _ode_ returns a matrix which then is 
converted to a data frame for convenience. readsdr offers **sd_simulate** to 
simplify this process.


```{r}
deSolve_components <- mdl$deSolve_components

all.equal(deSolve_components, xmile_to_deSolve(filepath))

library(deSolve)

simtime <- seq(deSolve_components$sim_params$start,
               deSolve_components$sim_params$stop,
               deSolve_components$sim_params$dt)

output_deSolve <- ode(y      = deSolve_components$stocks,
                      times  = simtime,
                      func   = deSolve_components$func,
                      parms  = deSolve_components$consts, 
                      method = "euler")

result_df <- data.frame(output_deSolve)

head(result_df)

result_df2 <- sd_simulate(deSolve_components)
identical(result_df, result_df2)
```

With the results in a data frame, we can manipulate the data structure for 
further analysis and data visualisation. In this case, we simply produce a plot
with the behaviour over time using ggplot2, a library part of the [tidyverse](https://www.tidyverse.org/).
See [Duggan (2018)](https://onlinelibrary.wiley.com/doi/abs/10.1002/sdr.1600) for
more details.

```{r, message = FALSE, warning = FALSE, fig.align='center', fig.width = 6}
library(dplyr)
library(tidyr)
library(ggplot2)

tidy_result_df <- result_df %>% 
  select(time, Susceptible, Infected, Recovered) %>%
  pivot_longer(-time, names_to = "Variable") 

ggplot(tidy_result_df, aes(x = time, y = value)) +
  geom_line(aes(group = Variable, colour = Variable)) +
  theme_classic() +
  theme(legend.position = "bottom")
```

Finally, the third element from _read_xmile_ is a list of two data frames. 
These structures describe the model as a graph in terms of vertices and edges,
and serve as input to the functions in [igraph](https://igraph.org/), a library 
collection for creating and manipulating graphs and analyzing networks.

```{r, fig.align='center', fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
library(igraph)

gr <- graph_from_data_frame(mdl$graph_dfs$edges, directed = TRUE, 
                               vertices = mdl$graph_dfs$nodes)

V(gr)$shape <- ifelse(V(gr)$type == "stock", "rectangle", "circle")

plot.igraph(gr, edge.arrow.size = 0.25,
            vertex.label.cex = 1, layout = layout.lgl(gr))
```

